\paragraph{Setup.}
Let
\[
\psi_q(\y;\bt)=m_q(\y;\bt)-c\,b_q(\y;\bt),\qquad
A_c(\bt_0)=\E[\nabla_\bt\psi_q(\Y;\bt)]_{\bt=\bt_0},\qquad
B_c(\bt_0)=\Var(\psi_q(\Y;\bt_0)),
\]
and denote the Fisher information by \(I\equiv I(\bt_0)=\Var(s(\bt_0;\Y))\succ0\).
As shown earlier,
\[
A_c(\bt_0)\;=\;-\,I^{1/2}\big(I+\widetilde M_c\big)I^{1/2},\qquad
\widetilde M_c
=I^{-1/2}\,\E\!\big[(m_q-c\,b_q-s)\,s^\top\big]\,I^{-1/2}.
\]

\begin{proposition}[Clean robustness–efficiency knob]
\label{prop:clean-knob}
Fix \(\bt_0\). Then:
\begin{enumerate}
\item[\textup{(E)}] \textbf{Efficiency at the oracle.} If \(q=f\) (true posterior),
\[
m_q(\cdot;\bt_0)-b_q(\cdot;\bt_0)=s(\bt_0;\cdot),\quad
\Rightarrow\quad
A_{1}(\bt_0)=-I,\ \ B_{1}(\bt_0)=I,\ \ \Sigma_{1}=I^{-1}.
\]
Moreover, \(A_c\to -I\) and \(B_c\to I\) as \(c\uparrow1\); hence \(\Sigma_c\to I^{-1}\).
\item[\textup{(R)}] \textbf{Robustness under misspecification.} For general \(q\),
write \(\widetilde M_c=\widetilde M^{(1)}-c\,\widetilde M^{(2)}\) with
\[
\widetilde M^{(1)}=I^{-1/2}\E[(m_q-s)\,s^\top]I^{-1/2},\qquad
\widetilde M^{(2)}=I^{-1/2}\E[b_q\,s^\top]I^{-1/2}.
\]
Then
\[
\|\widetilde M_c\|_{\op}\ \le\ \|\widetilde M^{(1)}\|_{\op}\ +\ c\,\|\widetilde M^{(2)}\|_{\op}.
\]
Hence the sufficient identifiability test
\(\ \|\widetilde M_c\|_{\op}<1\) is \emph{easiest to satisfy at \(c=0\)}.
\end{enumerate}
\end{proposition}

\begin{proof}[Proof (one–liners)]
\textup{(E)}: With \(q=f\), \(m_q-b_q=s\) and thus \(\widetilde M_{1}=0\).
So \(A_{1}=-I^{1/2} I I^{1/2}=-I\), \(B_{1}=\Var(s)=I\), whence \(\Sigma_{1}=I^{-1}\).
Continuity of \(c\mapsto A_c,B_c\) yields \(\Sigma_c\to I^{-1}\) as \(c\uparrow1\).

\textup{(R)}: Decompose \(\widetilde M_c\) as stated and apply the triangle inequality.
Since \(A_c=-I^{1/2}(I+\widetilde M_c)I^{1/2}\), the Neumann condition
\(\|\widetilde M_c\|_{\op}<1\) guarantees \(A_c\) invertible; the bound is minimized at \(c=0\).
\end{proof}

\paragraph{Intuition (one paragraph).}
The term \(b_q\) is a model–based centering computed through the surrogate posterior.
At \(c=1\) and a good encoder (\(q\approx f\)), it cancels the bias and recovers the score,
giving MLE efficiency. When \(q\) is misspecified, \(b_q\) can inject error into the Jacobian.
Reducing \(c\) down-weights reliance on \(b_q\), shrinking the perturbation of \(A_c\) and
making invertibility easiest to secure. Thus \(c=0\) is most \emph{robust}; \(c=1\) is most
\emph{efficient} when the encoder is accurate.
